\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-09-18 analysis/manuscript.tex
Recompile tex
Minor fixes and recompilation of manuscript

% bibliography
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{CSLReferences}%
  {}%
  {\par}

@@

\abstract{The ongoing controversy surrounding transformative agreements, which aim to transition journal publishing to full open access, highlight the need for large-scale studies assessing the impact of these agreements on hybrid open access. By combining publicly available data from various sources, including cOAlition S Journal Checker, Crossref, and OpenAlex, this study presents a novel approach that analyses over 700 agreements. Results suggest a strong growth in open access between 2018 and 2022 from 4.3\% to 15\%. During this period, 11,189 hybrid journals provided open access to 742,369 out of 8,146,958 articles, representing a five-year open access proportion of 9.1\%. Authors who could make use of transformative agreements at the time of publication contributed 328,957 open access articles. In 2022, 143,615 out of 249,511 open access articles in hybrid journals (58\%) were enabled by transformative agreements. This trend was largely driven by the three commercial publishers Elsevier, Springer Nature, and Wiley, but the open access uptake varied substantially across journals, publishers, disciplines, and country affiliations. In particular, the OECD and BRICS areas revealed different publication trends. In conclusion, this study suggests that current levels of implementation of transformative agreements is insufficient to bring about a large-scale transition to full open access.}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing (Mittermaier, 2015), as well as to increase
publisher-provided immediate open access (Björk \& Solomon, 2014;
Schimmer et al., 2015). These considerations resulted in transformative
agreements\footnote{In this paper I use the term ``transformative
  agreement'', addressing also offsetting, read-and-publish or
  publish-and-read deals, and other variants (Borrego et al., 2021;
  Hinchliffe, 2019). Although the term is critised as misleading and not
  useful to describe the different types of open access agreements
  between library consortia and commercial publishers (Babini et al.,
  2022), it is widely used in policy discussions and in the research
  literature.}, which cover a broad range of contracts between library
consortia and publishers from the mid-2010s onwards, where institutional
spending for subscriptions and open access publishing are considered
together (Borrego et al., 2021; Hinchliffe, 2019). Transformative
agreements seek to control costs while allowing a transitional phase for
publishing more open access articles. Similar to big deals,
transformative agreements mainly bundle hybrid and subscription-only
journals from commercial publishers, but aim at a higher degree of
transparency than previous big deals, where contracts including payments
were confidential (Bergstrom et al., 2014).

@@

By the end of 2023, many transformative agreements had been implemented,
but the interim outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest resource for library consortia to publicly share their
agreements, recorded more than 800 transformative agreements. These
agreements resulted in the publication of up to 900.000 open access
articles published in both fully open access and hybrid journals,
according to the accompanying ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilisation of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements has contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuates market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Ross-Hellauer et al., 2022), because transformative
agreements' focus on pay to publish mainly targets institutions from
high-income countries, furthering a questionable journal prestige
culture (Babini et al., 2022). Besides, an editorial-board resignation
raised concerns that transformative agreements might encourage
publishers to maximize journal publication volume ``without regard to
quality'' (Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended only supporting
agreements for publishing in fully open access journals by 2026
\footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023).
cOAlition S also removed the majority of hybrid journals from its
Transformative Journal program in 2023 due to publishers' failure to
meet self-defined open access growth targets (Brainard, 2023). By
contrast, the German DEAL consortium announced a five-year
transformative agreement with Elsevier starting in 2024 and also renewed
its contracts with Springer Nature and Wiley until the end of 2028.
Similarly, the Colombia Consortium signed the first transformative
agreements in Latin America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the uptake of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Brayman et al., 2024; Haucap et al., 2021; Huang et al.,
2020; Pölönen et al., 2020; Taubert et al., 2023; Wenaas, 2022) or
publisher portfolios (Bakker et al., 2024; Fraser et al., 2023; Jahn et
al., 2022; Momeni et al., 2023; Pieper \& Broschinski, 2018; Schmal,
2024), while large-scale studies relied on self-reported agreement data
(Moskovkin et al., 2022), or used APC pricing lists (Shu \& Larivière,
2023). In particular, data availability is a limiting factor when
studying the impact of transformative agreements (Bakker et al., 2024),
because bibliometric databases, even though many allow the retrieval of
open access articles in hybrid journals, do not directly attribute them
to specific transformative agreements. Likewise, article-level open
access invoicing and cost data are only partly available (Jahn et al.,
2022; Kramer, 2024).

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref (Hendricks et al., 2020) served as the primary
data source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. To determine open access articles published through
transformative agreements, first author affiliations from OpenAlex
(Priem et al., 2022) were subsequently linked to eligible institutions
according to the transformative agreement data. In the following, the
steps are described in detail.

@@

\hypertarget{transformative-agreement-and-hybrid-journal-data}{%
\subsection{Transformative agreement and hybrid journal
data}\label{transformative-agreement-and-hybrid-journal-data}}

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool.\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
The dump consists of multiple online Google spreadsheets where each data
file represents one agreement listed in the ESAC Transformative
Agreement Registry.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}
From the retrieved spreadsheet files, journals and institutions involved
per agreement were obtained.

A limitation of using the Journal Checker Tool and its underlying
publicly available data dump to study the development of transformative
agreements over time is that expired transformative agreements are
constantly removed. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included, representing
the majority of transformative agreements. Overall, the four combined
Transformative Agreement Data dumps used in this study contained 729 of
869 agreements listed in the ESAC Transformative Agreement Registry by
December 2023.

@@

\hypertarget{article-and-author-metadata}{%
\subsection{Article and author
metadata}\label{article-and-author-metadata}}

@@

Open access articles in hybrid journals were identified using the
Creative Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version was not
considered. Crossref was used for open access identification because
transformative agreement workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017).

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements because they are
considered responsible for arranging open access publication (Borrego et
al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). For this
study, country and institutional affiliations were retrieved from
OpenAlex. However, because corresponding author affiliation was only
available for 54\% of the investigated articles in OpenAlex, this study
focused on first authors and their affiliations instead; approximately
90\% of the investigated articles had first author affiliation metadata
in OpenAlex, which is a much larger proportion than previously reported
for the October 2022 snapshot (Zhang et al., 2024). First authors
typically contribute the most to a paper and are often considered being
lead authors (Larivière et al., 2016). Related studies also assumed
first authors as a proxy to examine open access payments and
transformative agreements (Haucap et al., 2021; Shu \& Larivière, 2023;
Zhang et al., 2022).

To estimate the impact of transformative agreements on hybrid open
access, participating institutions from the Transformative Agreement
Data dump, which the cOAlition S crowd-sourced from agreement documents
and consortia, were matched with the first author affiliations recorded
by OpenAlex using the ROR-ID. Matching also considered the duration of
agreements according to the ESAC registry. Upon inspection,
Transformative Agreement Data did not comprehensively cover associated
institutions, such as university hospitals or institutes of large
research organisations such as the Max Planck Society. To improve the
matching, Transformative Agreement Data were automatically enriched with
associated organisations using OpenAlex's institution entity data.

The so compiled data set consists of 8,922,146 articles published in
12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals in transformative agreements
represented 40\% of the total global output over the same time period,
according to Crossref, while fully open access journals recorded 35\%.

@@

\caption{Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The blue bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comparision with full open access journals according to OpenAlex. The remainder represents subscription-based journals not covered by transformative agreements.}\label{fig:method_fig}
\end{figure}

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{overview}{%
\subsection{Overview}\label{overview}}

Between 2018 and 2022, 11,189 out of 12,857 hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 out of 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles. Overall, 394 transformative
agreements recorded at least one open access article from an eligible
institution.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). Simultaneously,
the total article volume of the investigated journals rose from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022. In 2022, 143,615 out of 249,511 open access articles
were from eligible authors, representing 58\%. However, there was also a
notable increase in open access provision through other means,
presumably publication fees being not invoiced through transformative
agreements, from 4.1\% (n = 62,625) in 2018 to 6.3\% (n = 105,896) in
2022.

@@

\hypertarget{publishing-market}{%
\subsection{Publishing market}\label{publishing-market}}

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of the total article
volume published (see Table \ref{tab:publisher_league_table}). Together,
they published 500,878 or 66\% of the open access articles in hybrid
journals. Elsevier, Springer Nature, and Wiley made 243,891 articles
open access in hybrid journals through transformative agreements,
resulting in an even larger market share of 74\%.

@@

\caption{\label{tab:publisher_league_table}Hybrid open access through transformative agreements market shares 2018-2022}
\centering
\begin{tabular}[t]{lrlrlrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hybrid journals} & \multicolumn{2}{c}{Articles} & \multicolumn{2}{c}{OA articles} & \multicolumn{2}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Publisher & Total & \% & Total & \% & Total & \% & Total & \%\\
\midrule
Elsevier & 1,936 & 17 & 2,770,826 & 33.8 & 172,723 & 22.9 & 60,440 & 18.3\\
Springer Nature & 2,274 & 20 & 1,330,430 & 16.2 & 175,432 & 23.3 & 100,008 & 30.3\\
Wiley & 1,410 & 12.4 & 1,043,052 & 12.7 & 152,723 & 20.3 & 83,443 & 25.3\\
Other & 5,767 & 50.6 & 3,061,337 & 37.3 & 252,523 & 33.5 & 86,294 & 26.1\\
\bottomrule
\end{tabular}
\end{table}

@@

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year, with a focus on open
access enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that Wiley experienced a
substantial increase in its open access share from 5.9\% (n = 11,628) in
2018 to 26\% (n = 53,503) in 2022, representing a 4.5-fold increase. In
contrast, Elsevier's hybrid journals demonstrated a more modest
increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n = 60,821) in 2022,
which is a relatively low open access share compared to the general
trend. In 2018, Springer Nature had the largest open access proportion
among the three publishers of 8.4\% (n = 19,701), but experienced a
relatively slower growth, resulting in 18\% (n = 52,616) of articles
being open access in Springer Nature hybrid journals in 2022.

@@

\caption{Developement of open access in hybrid journals included in transformative agreements between 2018 and 2022 by publishers. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:publisher_figure}
\end{figure}

>>>>

% bibliography
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
  {\begin{list}{}{%
   \setlength{\itemindent}{0pt}
   \setlength{\leftmargin}{0pt}
   \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
    \setlength{\leftmargin}{\cslhangindent}
    \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
   \setlength{\itemsep}{#2\baselineskip}}}
  {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\abstract{The ongoing controversy surrounding transformative agreements, which aim to transition subscription-based journal publishing to full open access, highlights the need for large-scale studies assessing the impact of these agreements on hybrid open access. By combining multiple open data sources, including cOAlition S Journal Checker, Crossref, and OpenAlex, this study presents a novel approach that analyses over 700 agreements. Results suggest a strong growth in open access, from 4.3\% in 2018 to 15\% in 2022. Over five years, 11,189 hybrid journals provided open access to 742,369 out of 8,146,958 articles (9.1\%). Authors who could make use of transformative agreements contributed 328,957 open access articles (44\%) during this period, reaching a peak in 2022 with 143,615 out of 249,511 open access articles (58\%). While this trend was predominantly driven by the three commercial publishers Elsevier, Springer Nature, and Wiley, open access uptake varied substantially across journals, publishers, disciplines, and countries. Particularly, the OECD and BRICS areas revealed different publication trends. In conclusion, this study suggests that current levels of implementation of transformative agreements is insufficient to bring about a large-scale transition to full open access.}

@@

\section{Introduction}\label{introduction}

@@

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing fees (Mittermaier, 2015), as well as to
increase publisher-provided immediate open access (Björk \& Solomon,
2014; Schimmer et al., 2015). These considerations resulted in
transformative agreements\footnote{In this paper I use the term
  ``transformative agreement'', addressing also offsetting,
  read-and-publish or publish-and-read deals, and other variants
  (Borrego et al., 2021; Hinchliffe, 2019). Although the term is
  critised as misleading and not useful to describe the different types
  of open access agreements between library consortia and commercial
  publishers (Babini et al., 2022), it is widely used in policy
  discussions and in the research literature.}, which cover a broad
range of contracts between library consortia and publishers from the
mid-2010s onwards, where institutional spending for subscriptions and
open access publishing are considered together (Borrego et al., 2021;
Hinchliffe, 2019). Transformative agreements seek to control costs while
allowing a transitional phase for publishing more open access articles.
Similar to big deals, transformative agreements mainly bundle hybrid and
subscription-only journals from commercial publishers, but aim at a
higher degree of transparency than previous big deals, where contracts
including payments were confidential (Bergstrom et al., 2014).

@@

By the end of 2023, many transformative agreements had been implemented,
but the interim outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest resource for library consortia to voluntarily and publicly
share their agreements, recorded more than 800 transformative
agreements. These agreements resulted in the publication of up to
900.000 open access articles published in both fully open access and
hybrid journals, according to the accompanying ESAC Market
Watch\footnote{\url{https://esac-initiative.org/market-watch/}}. Library
consortia reported increased open access volume, streamlined payment and
monitoring procedures, as well as extensive utilisation of open access
options by the researchers they serve (Marques \& Stone, 2020; Parmhed
\& Säll, 2023; Pinhasi et al., 2020). The ongoing standardisation of
transformative agreements has contributed to improved transparency in
terms of contracts and publisher-provided article metadata (Marques et
al., 2019; Pinhasi et al., 2021). However, with the growing trend toward
transformative agreements, continued reliance on big deals is perceived
as problematic, because it perpetuates market concentration (Butler et
al., 2023; Shu \& Larivière, 2023). Whether transformative agreements
lead to reduced pricing remains uncertain (Borrego, 2023) and a
substantial transition of hybrid journals towards full open access could
not be observed (Matthias et al., 2019; Momeni et al., 2021). The focus
on large commercial publishers might also increase inequality
(Ross-Hellauer et al., 2022), because transformative agreements' focus
on pay to publish mainly targets institutions from high-income
countries, furthering a questionable journal prestige culture (Babini et
al., 2022). Besides, an editorial-board resignation raised concerns that
transformative agreements might encourage publishers to maximize journal
publication volume ``without regard to quality'' (Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
UK's Joint Information Systems Committee (JISC) evaluated its open
access strategy, which included transformative agreements (Brayman et
al., 2024). The evaluation revealed that while these agreements had a
significant impact on the growth of open access in the UK, they had
limited effects in facilitating a global shift towards full open access.
As a result, the report advised that the UK's open access strategy
should be reassessed. After making similar observation, Norwegian and
Swedish universities and their consortia also argued for policy changes
(Holden et al., 2023; Widding, 2024). Notably, the Association of
Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended only supporting agreements for
publishing in fully open access journals by 2026 (Widding, 2024).
Furthermore, cOAlition S will conclude its financial support of
transformative agreements at the end of 2024, but will continue to view
open access resulting from such agreements as compliant. \footnote{\url{https://www.coalition-s.org/coalition-s-confirms-the-end-of-its-financial-support-for-open-access-publishing-under-transformative-arrangements-after-2024/}}.
cOAlition S also removed the majority of hybrid journals from its
Transformative Journal program in 2023 due to publishers' failure to
meet self-defined open access growth targets (Brainard, 2023). By
contrast, the German DEAL consortium announced a five-year
transformative agreement with Elsevier starting in 2024 and also renewed
its contracts with Springer Nature and Wiley until the end of 2028.
Similarly, the Colombia Consortium signed the first transformative
agreements in Latin America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the uptake of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Brayman et al., 2024; Haucap et al., 2021; Huang et al.,
2020; Pölönen et al., 2020; Taubert et al., 2023; Waltman \& Lamers,
2022; Wenaas, 2022) or publisher portfolios (Bakker et al., 2024; Fraser
et al., 2023; Jahn et al., 2022; Momeni et al., 2023; Pieper \&
Broschinski, 2018; Schmal, 2024), while large-scale studies relied on
self-reported agreement data (Moskovkin et al., 2022), or used APC
pricing lists (Shu \& Larivière, 2023). In particular, data availability
is a limiting factor when studying the impact of transformative
agreements (Bakker et al., 2024), because bibliometric databases, even
though many allow the retrieval of open access articles in hybrid
journals, do not directly attribute them to specific transformative
agreements. Likewise, article-level open access invoicing and cost data
are only partly available, which would make it possible to establish a
direct link between transformative agreements and open access
publications (Jahn et al., 2022; Kramer, 2024).

@@

\section{Methods}\label{methods}

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref (Hendricks et al., 2020) served as the primary
data source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. Because of a lack of comprehensive publicly available
invoicing data, open access articles published through transformative
agreements were determined by first author affiliations from OpenAlex
(Priem et al., 2022), linked to eligible institutions according to the
transformative agreement data. In the following, the steps are described
in detail.

@@

\subsection{Transformative agreement and hybrid journal
data}\label{transformative-agreement-and-hybrid-journal-data}

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool \footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}},
a voluntary effort based on publicly disclosed contracts in the ESAC
Transformative Agreement Registry.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.
The dump consists of multiple online Google spreadsheets where each data
file represents one agreement listed in the ESAC Transformative
Agreement Registry. From the retrieved spreadsheet files, journals and
institutions involved per agreement were obtained.

It should be noted that although many library consortia see the need to
register their agreements through the ESAC registry, many fail to
disclose all details, including contracts (Bakker et al., 2024; Kramer,
2024). Furthermore, the extent to which ESAC is comprehensive remains
uncertain, potentially limiting the coverage of transformative
agreements in the Journal Checker Tool. Another limitation of using the
Journal Checker Tool and its underlying publicly available data dump to
study the development of transformative agreements over time is that
expired transformative agreements are constantly removed. To address
this, four different snapshots were safeguarded and combined for this
study: self-archived versions from July 2021, July 2022, and May 2023,
as well as the most current dump downloaded on 11 December 2023. This
ensured that transformative agreements, which ended from 2021 onwards,
were included, representing the majority of transformative agreements.
Overall, the four combined Transformative Agreement Data dumps used in
this study contained 729 of 869 agreements listed in the ESAC
Transformative Agreement Registry by December 2023.

@@

\subsection{Article and author
metadata}\label{article-and-author-metadata}

@@

Open access articles in hybrid journals were identified using the
Creative Commons (CC) license information in Crossref metadata, with
consideration of any CC variant. License information relative to the
``accepted manuscript (AM)'' version was not considered. Crossref was
used for open access identification because transformative agreement
workflows generally require publishers to deliver CC license information
to this DOI registration agency (Geschuhn \& Stone, 2017).

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements because they are
considered responsible for arranging open access publication (Borrego et
al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). For this
study, country and institutional affiliations were retrieved from
OpenAlex. Because the corresponding authorship field was not fully
supported by OpenAlex at the time of analysis, and respective
affiliation data were only available for 54\% of the investigated
articles, this study focused on first authors and their affiliations
instead; approximately 90\% of the articles examined had first author
affiliation metadata in OpenAlex, which is a much larger proportion than
previously reported for the October 2022 snapshot (Zhang et al., 2024).
First authors typically contribute the most to a paper and are often
considered lead authors (Larivière et al., 2016), although notable
differences in corresponding author metadata were observed across
subjects, publication years and bibliometric databases
(Chinchilla-Rodríguez et al., 2024). In case of lacking data about
corresponding authors in bibliometric databases, related studies also
utilised first authors as a proxy to examine open access payments and
transformative agreements (Haucap et al., 2021; Shu \& Larivière, 2023;
Zhang et al., 2022).

To estimate the impact of transformative agreements on hybrid open
access, participating institutions from the Transformative Agreement
Data dump, which the cOAlition S crowd-sourced from agreement documents
and consortia, were matched with the first author affiliations recorded
by OpenAlex using the ROR-ID. Matching also considered the duration of
agreements according to the ESAC registry. 502 agreements were active
between 2018 and 2022. Upon inspection, Transformative Agreement Data
did not comprehensively cover associated institutions, such as
university hospitals or institutes of large research organisations such
as the Max Planck Society. To improve the matching, Transformative
Agreement Data were automatically enriched with ROR-IDs from associated
organisations according to OpenAlex's institution entity data.

@@

\caption{Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The purple bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comparision with full open access journals according to OpenAlex. The remainder represents articles in subscription-based journals not covered by transformative agreements.}\label{fig:method_fig}
\end{figure}

The so compiled data set consists of 8,922,146 articles published in
12,857 hybrid journals included in at least one transformative agreement
between 2018 and 2022 (see purple bars in Figure \ref{fig:method_fig}B).
These hybrid journals in transformative agreements represented 40\% of
the total global output over the same time period, according to
Crossref, while fully open access journals recorded 35\%.

\subsection{Data analysis}\label{data-analysis}

@@

\section{Results}\label{results}

\subsection{Overview}\label{overview}

Between 2018 and 2022, 11,189 out of 12,857 eligible hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. These 11,189 hybrid journals
constituted the foundation for the subsequent analyses. Collectively,
they provided open access to 742,369 out of 8,146,958 articles during
the investigated period, representing a five-year open access proportion
of 9.1\%. Authors who could make use of transformative agreements at the
time of publication contributed 328,957 (44\%) open access articles.
Overall, this investigation was able to establish a link between open
access articles and eligible institutions for 394 out of 502 (78\%)
active transformative agreements between 2018 and 2022 through the
analysis of publication metadata, including affiliations and agreement
data.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
in hybrid journals increased from 4.3\% (n = 65,486) to 15\% (n =
249,511). Simultaneously, the total article volume of the investigated
journals rose from 1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022. In 2022, 143,615 out of 249,511 open access articles
were from eligible authors, representing 58\%. However, there was also a
notable increase in open access provision through other means,
presumably publication fees being not invoiced through transformative
agreements as indicated by the grey area in \ref{fig:results_overview}A,
from 4.1\% (n = 62,625) in 2018 to 6.3\% (n = 105,896) in 2022.

@@

\subsection{Publishing market}\label{publishing-market}

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the three commercial publishers Elsevier,
Springer Nature, and Wiley accounted for 49\% of hybrid journals,
representing 5,144,308 or 63\% of the total article volume (see Table
\ref{tab:publisher_league_table}). Together, they published 500,878 or
66\% of the open access articles in hybrid journals. Elsevier, Springer
Nature, and Wiley made 243,891 articles open access in hybrid journals
through transformative agreements, resulting in an even larger market
share of 74\%.

@@

\caption{\label{tab:publisher_league_table}Hybrid open access through transformative agreements market shares 2018-2022}
\centering
\begin{tabular}[t]{lrlrlrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hybrid journals} & \multicolumn{2}{c}{Articles} & \multicolumn{2}{c}{OA Articles} & \multicolumn{2}{c}{TA OA Articles} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Publisher & Total & \% & Total & \% & Total & \% & Total & \%\\
\midrule
Elsevier & 1,936 & 17 & 2,770,826 & 33.8 & 172,723 & 22.9 & 60,440 & 18.3\\
Springer Nature & 2,274 & 20 & 1,330,430 & 16.2 & 175,432 & 23.3 & 100,008 & 30.3\\
Wiley & 1,410 & 12.4 & 1,043,052 & 12.7 & 152,723 & 20.3 & 83,443 & 25.3\\
Other & 5,767 & 50.6 & 3,061,337 & 37.3 & 252,523 & 33.5 & 86,294 & 26.1\\
\bottomrule
\end{tabular}
\end{table}

@@

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year, with a focus on open
access enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that Wiley experienced a
substantial increase in its open access share from 5.9\% (n = 11,628) in
2018 to 26\% (n = 53,503) in 2022. In contrast, Elsevier's hybrid
journals demonstrated a more modest increase, from 3.3\% (n = 16,872) in
2018 to 10\% (n = 60,821) in 2022, which is a relatively low open access
share compared to the general trend. In 2018, Springer Nature had the
largest open access proportion among the three publishers of 8.4\% (n =
19,701), but experienced a relatively slower growth, resulting in 18\%
(n = 52,616) of articles being open access in Springer Nature hybrid
journals in 2022.

@@

\caption{Development of open access in hybrid journals included in transformative agreements between 2018 and 2022 by publishers. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:publisher_figure}
\end{figure}

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, started offering open access
agreements for its hybrid journal portfolio to selected consortia such
as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish
FinELib consortium in 2015 under the name Springer Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
Prior to this, Springer had some pilot agreements with a small number of
institutions, including the University of Göttingen (Schmidt \& Shearer,
2012). However, the Springer Compact agreements were not included in the
data as they concluded before the start of the transformative agreement
data collection in June 2021. Nonetheless, the results demonstrate the
importance of agreements for Springer Nature's hybrid open access
business over the past five years (Figure \ref{fig:publisher_figure}).
In 2022, 66\% (n = 34,725) of open access in Springer Nature hybrid
journals was enabled through transformative agreements. In the same
year, 70\% (n = 37,316) of Wiley's open access articles could be linked
to transformative agreements. By contrast, Elsevier published a
comparatively lower proportion of its open access articles through
transformative agreements in 2022 (n = 32,627; 54\%).
